\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-10-26 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd

In this article, we present the \CRANpkg{casebase} R package [@casebase-package] 
which implements the Hanley & Miettinen [-@hanley2009fitting] approach for 
fitting fully parametric hazard models and covariate-adjusted survival curves 
using the familiar interface of the \code{glm} function. Our implementation 
allows for straightforward extensions to other models such as penalized 
regression for variable selection and competing-risk analysis. In addition, 
we provide functions for exploratory data analysis and visualizing the estimated 
quantities such as the hazard function, survival curve, and their standard errors. 
The ultimate goal of our package is to make fitting flexible hazards accessible 
to more end users with the hope that they will favor reporting absolute risks 
over hazard ratios.

@@

As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to consider the entire study base as an infinite collection of *person moments*. These person moments are indexed by both an individual in the study and a time point, and therefore each person moment has a covariate profile, an exposure status, and an outcome status (i.e. whether the event happened) attached to it. By comparing person moments at which an event occurred with person moments at which no event occurred, we can extract information about hazards and survival.

Therefore, we start by sampling all person moments at which the event occurred; this collection of person moments is what Hanley & Miettinen call the *case series*. The incidence of the case series is dictated by the hazard function of interest. Next, we sample a finite number of person moments at which no event occurred; this second collection of person moment is what Hanley & Miettinen call the *base series*. The sampling mechanism for the base series is left at the discretion of the user, but in practice we find that sampling uniformly from the study base provides both simplicity and good performance. This is the default sampling mechanism in the package.

@@

To describe the theoretical foundations of case-base sampling, we use the framework of counting processes. In what follows, we abuse notation slightly and omit any mention of $\sigma$-algebras. Instead, following Aalen \textit{et al} [-@aalen2008survival], we use the placeholder "past" to denote the past history of the corresponding process. The reader interested in more details can refer to Saarela & Arjas [-@saarela2015non] and Saarela [-@saarela2016case]. First, let $N_{i}(t) \in \{0, 1\}$ be counting processes corresponding to the event of interest for individual $i=1, \ldots,n$. For simplicity, we will consider Type I censoring due to the end of follow-up at time $\tau$ (the general case of non-informative censoring is treated in Saarela [-@saarela2016case]). We assume a continuous time model, which implies that the counting process jumps are less than or equal to one. We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
$$\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].$$  The processes $N_i(t)$ count the person moments from the *case series*.

To complement the case series, we sample person moments for the base series. To do so, we model the base series sampling mechanism using non-homogeneous Poisson processes $R_i(t) \in \{0, 1, 2, \ldots\}$; the person-moments where $dR_i(t) = 1$ constitute the base series. We note that the same individual can contribute multiple person-moments to the base series. The process $Q_{i}(t) = R_i(t) + N_{i}(t)$ then counts both the case and base series person-moments contributed by individual $i$. As mentioned above, the processes $R_i(t)$ are typically defined by the user via its intensity function $\rho_i(t)$. The process $Q_{i}(t)$ is characterized by $E[dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)dt + \rho_i(t)dt$.

>>>>

In this article, we present the \CRANpkg{casebase} R package [@casebase-package] 
which implements and extends the Hanley & Miettinen [-@hanley2009fitting] approach for 
fitting fully parametric hazard models and covariate-adjusted survival curves 
using the familiar interface of the \code{glm} function. Our implementation 
includes extensions to other models such as penalized 
regression for variable selection and competing-risk analysis. In addition, 
we provide functions for exploratory data analysis and visualizing the estimated 
quantities such as the hazard function, survival curve, and their standard errors. 
The ultimate goal of our package is to make fitting flexible hazards accessible 
to more end users with the hope that they will favor reporting absolute risks and survival curves
over hazard ratios.

@@

As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to consider the entire study base as an infinite collection of *person-moments*. These person-moments are indexed by both an individual in the study and a time point, and therefore each person-moment has a covariate profile, an exposure status, and an outcome status (i.e. whether the event happened) attached to it. By comparing person-moments at which an event occurred with person-moments at which no event occurred, we can extract information about hazards and survival.

Therefore, we start by sampling all person-moments at which the event occurred; this collection of person-moments is what Hanley & Miettinen call the *case series*. The incidence of the case series is dictated by the hazard function of interest. Next, we sample a finite number of person-moments at which no event occurred; this second collection of person-moment is what Hanley & Miettinen call the *base series*. The sampling mechanism for the base series is left at the discretion of the user, but in practice we find that sampling uniformly from the study base provides both simplicity and good performance. This is the default sampling mechanism in the package.

@@

To describe the theoretical foundations of case-base sampling, we use the framework of counting processes. In what follows, we abuse notation slightly and omit any mention of $\sigma$-algebras. Instead, following Aalen \textit{et al} [-@aalen2008survival], we use the placeholder "past" to denote the past history of the corresponding process. The reader interested in more details can refer to Saarela & Arjas [-@saarela2015non] and Saarela [-@saarela2016case]. First, let $N_{i}(t) \in \{0, 1\}$ be counting processes corresponding to the event of interest for individual $i=1, \ldots,n$. For simplicity, we will consider Type I censoring due to the end of follow-up at time $\tau$ (the general case of non-informative censoring is treated in Saarela [-@saarela2016case]). We assume a continuous time model, which implies that the counting process jumps are less than or equal to one. We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
$$\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].$$  The processes $N_i(t)$ count the person-moments from the *case series*.

To complement the case series, we sample person-moments for the base series. To do so, we model the base series sampling mechanism using non-homogeneous Poisson processes $R_i(t) \in \{0, 1, 2, \ldots\}$; the person-moments where $dR_i(t) = 1$ constitute the base series. We note that the same individual can contribute multiple person-moments to the base series. The process $Q_{i}(t) = R_i(t) + N_{i}(t)$ then counts both the case and base series person-moments contributed by individual $i$. As mentioned above, the processes $R_i(t)$ are typically defined by the user via its intensity function $\rho_i(t)$. The process $Q_{i}(t)$ is characterized by $E[dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)dt + \rho_i(t)dt$.

### 2021-10-26 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex

In this article, we present the \CRANpkg{casebase} R package
\citep{casebase-package} which implements the Hanley \& Miettinen
\citeyearpar{hanley2009fitting} approach for fitting fully parametric
hazard models and covariate-adjusted survival curves using the familiar
interface of the \code{glm} function. Our implementation allows for
straightforward extensions to other models such as penalized regression
for variable selection and competing-risk analysis. In addition, we
provide functions for exploratory data analysis and visualizing the
estimated quantities such as the hazard function, survival curve, and
their standard errors. The ultimate goal of our package is to make
fitting flexible hazards accessible to more end users with the hope that
they will favor reporting absolute risks over hazard ratios.

@@

As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
key idea behind case-base sampling is to consider the entire study base
as an infinite collection of \emph{person moments}. These person moments
are indexed by both an individual in the study and a time point, and
therefore each person moment has a covariate profile, an exposure
status, and an outcome status (i.e.~whether the event happened) attached
to it. By comparing person moments at which an event occurred with
person moments at which no event occurred, we can extract information
about hazards and survival.

Therefore, we start by sampling all person moments at which the event
occurred; this collection of person moments is what Hanley \& Miettinen
call the \emph{case series}. The incidence of the case series is
dictated by the hazard function of interest. Next, we sample a finite
number of person moments at which no event occurred; this second
collection of person moment is what Hanley \& Miettinen call the
\emph{base series}. The sampling mechanism for the base series is left
at the discretion of the user, but in practice we find that sampling
uniformly from the study base provides both simplicity and good
performance. This is the default sampling mechanism in the package.

@@

To describe the theoretical foundations of case-base sampling, we use
the framework of counting processes. In what follows, we abuse notation
slightly and omit any mention of \(\sigma\)-algebras. Instead, following
Aalen \textit{et al} \citeyearpar{aalen2008survival}, we use the
placeholder ``past'' to denote the past history of the corresponding
process. The reader interested in more details can refer to Saarela \&
Arjas \citeyearpar{saarela2015non} and Saarela
\citeyearpar{saarela2016case}. First, let \(N_{i}(t) \in \{0, 1\}\) be
counting processes corresponding to the event of interest for individual
\(i=1, \ldots,n\). For simplicity, we will consider Type I censoring due
to the end of follow-up at time \(\tau\) (the general case of
non-informative censoring is treated in Saarela
\citeyearpar{saarela2016case}). We assume a continuous time model, which
implies that the counting process jumps are less than or equal to one.
We are interested in modeling the hazard functions \(\lambda_{i}(t)\) of
the processes \(N_i(t)\), and which satisfy
\[\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].\] The processes
\(N_i(t)\) count the person moments from the \emph{case series}.

To complement the case series, we sample person moments for the base
series. To do so, we model the base series sampling mechanism using
non-homogeneous Poisson processes \(R_i(t) \in \{0, 1, 2, \ldots\}\);
the person-moments where \(dR_i(t) = 1\) constitute the base series. We
note that the same individual can contribute multiple person-moments to
the base series. The process \(Q_{i}(t) = R_i(t) + N_{i}(t)\) then
counts both the case and base series person-moments contributed by
individual \(i\). As mentioned above, the processes \(R_i(t)\) are
typically defined by the user via its intensity function \(\rho_i(t)\).
The process \(Q_{i}(t)\) is characterized by
\(E[dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)dt + \rho_i(t)dt\).

>>>>

In this article, we present the \CRANpkg{casebase} R package
\citep{casebase-package} which implements and extends the Hanley \&
Miettinen \citeyearpar{hanley2009fitting} approach for fitting fully
parametric hazard models and covariate-adjusted survival curves using
the familiar interface of the \code{glm} function. Our implementation
includes extensions to other models such as penalized regression for
variable selection and competing-risk analysis. In addition, we provide
functions for exploratory data analysis and visualizing the estimated
quantities such as the hazard function, survival curve, and their
standard errors. The ultimate goal of our package is to make fitting
flexible hazards accessible to more end users with the hope that they
will favor reporting absolute risks and survival curves over hazard
ratios.

@@

As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
key idea behind case-base sampling is to consider the entire study base
as an infinite collection of \emph{person-moments}. These person-moments
are indexed by both an individual in the study and a time point, and
therefore each person-moment has a covariate profile, an exposure
status, and an outcome status (i.e.~whether the event happened) attached
to it. By comparing person-moments at which an event occurred with
person-moments at which no event occurred, we can extract information
about hazards and survival.

Therefore, we start by sampling all person-moments at which the event
occurred; this collection of person-moments is what Hanley \& Miettinen
call the \emph{case series}. The incidence of the case series is
dictated by the hazard function of interest. Next, we sample a finite
number of person-moments at which no event occurred; this second
collection of person-moment is what Hanley \& Miettinen call the
\emph{base series}. The sampling mechanism for the base series is left
at the discretion of the user, but in practice we find that sampling
uniformly from the study base provides both simplicity and good
performance. This is the default sampling mechanism in the package.

@@

To describe the theoretical foundations of case-base sampling, we use
the framework of counting processes. In what follows, we abuse notation
slightly and omit any mention of \(\sigma\)-algebras. Instead, following
Aalen \textit{et al} \citeyearpar{aalen2008survival}, we use the
placeholder ``past'' to denote the past history of the corresponding
process. The reader interested in more details can refer to Saarela \&
Arjas \citeyearpar{saarela2015non} and Saarela
\citeyearpar{saarela2016case}. First, let \(N_{i}(t) \in \{0, 1\}\) be
counting processes corresponding to the event of interest for individual
\(i=1, \ldots,n\). For simplicity, we will consider Type I censoring due
to the end of follow-up at time \(\tau\) (the general case of
non-informative censoring is treated in Saarela
\citeyearpar{saarela2016case}). We assume a continuous time model, which
implies that the counting process jumps are less than or equal to one.
We are interested in modeling the hazard functions \(\lambda_{i}(t)\) of
the processes \(N_i(t)\), and which satisfy
\[\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].\] The processes
\(N_i(t)\) count the person-moments from the \emph{case series}.

To complement the case series, we sample person-moments for the base
series. To do so, we model the base series sampling mechanism using
non-homogeneous Poisson processes \(R_i(t) \in \{0, 1, 2, \ldots\}\);
the person-moments where \(dR_i(t) = 1\) constitute the base series. We
note that the same individual can contribute multiple person-moments to
the base series. The process \(Q_{i}(t) = R_i(t) + N_{i}(t)\) then
counts both the case and base series person-moments contributed by
individual \(i\). As mentioned above, the processes \(R_i(t)\) are
typically defined by the user via its intensity function \(\rho_i(t)\).
The process \(Q_{i}(t)\) is characterized by
\(E[dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)dt + \rho_i(t)dt\).
